---
title: "Reading instances and parity with PyVRP"
description: >
  Read standard VRPLIB and Solomon instance files, and how vrpr compares to PyVRP.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.4,
  fig.align = "center",
  dpi = 120,
  out.width = "85%"
)
```

```{r}
#| label: setup
#| message: false
library(vrpr)
```

## Standard instance formats

`vrpr` ships native readers for the two most common benchmark formats, with no
Python dependency. Both return a `vrpr_model` you can solve right away.

### VRPLIB / TSPLIB

`read_vrplib()` reads CVRP (and VRPTW) instances in the VRPLIB/CVRPLIB format --
for example the well-known X set of Uchoa et al. It supports Euclidean
coordinates (`EDGE_WEIGHT_TYPE : EUC_2D`) and reads time-window and service-time
sections when present.

```{r}
#| label: vrplib
path <- system.file("extdata", "sample-n6-k2.vrp", package = "vrpr")

model <- read_vrplib(path)
model

res <- vrp_solve(model, stop = max_iterations(300), seed = 1, display = FALSE)
plot(res)
```

The number of vehicles is taken from the `VEHICLES`/`TRUCKS` field, or from the
`-k<n>` suffix in the instance name, or — as a last resort — the number of
clients. You can always override it with `num_vehicles =`.

### Solomon

`read_solomon()` reads VRPTW instances in the Solomon (and Gehring-Homberger)
format: a `VEHICLE` section with the fleet size and capacity, and a `CUSTOMER`
table with coordinates, demand, time window and service time. Customer 0 is the
depot.

```{r}
#| label: solomon
path <- system.file("extdata", "sample-solomon.txt", package = "vrpr")
read_solomon(path) |>
  vrp_solve(stop = max_iterations(300), seed = 1, display = FALSE) |>
  cost()
```

## Distances

When you do not supply explicit matrices, `vrpr` computes the Euclidean distance
between coordinates and rounds it half up (`floor(d + 0.5)`) — the TSPLIB
`EUC_2D` convention, which is what the published best-known solutions assume. You
can pass your own `distance` and `duration` matrices to `vrp_problem_data()` (and
hence to `vrp_solve()`) for road networks or asymmetric travel times.

## Parity with PyVRP

Because `vrpr` vendors PyVRP's C++ core, it is faithful to the reference solver
on two levels.

**Objective parity (exact).** The same solution scores identically on both sides,
since they share the same C++ `CostEvaluator`. On the bundled `sample-n6-k2`
instance, the routes `[[1, 2], [3, 4, 5]]` cost 81 in both `vrpr` and PyVRP.

**Quality parity.** On the `X-n101-k25` instance (100 clients, known optimum
**27591**), a 10-second release build of `vrpr` reaches the optimum, matching
PyVRP, with a comparable number of iterations. The reproducible benchmark lives
in `tools/benchmark/` (`parity.R` drives `vrpr`, `pyvrp_side.py` the reference).

| Solver | cost | gap to optimum |
|---|---|---|
| PyVRP | 27591 | 0.00 % |
| vrpr  | 27591 | 0.00 % |

> When benchmarking, always measure a **release** build (`R CMD INSTALL`), not the
> debug build of `devtools::load_all()`: the debug build compiles without
> optimisation and with the core's assertions enabled, making it many times
> slower and distorting any throughput comparison.

## The algorithm

`vrpr` runs the same iterated local search as PyVRP: an initial descent, then
repeated rounds of perturbation and local search, with Late Acceptance
Hill-Climbing (Burke & Bykov, 2017) as the acceptance criterion, a restart from
the best after a stretch without improvement, and an exhaustive refinement of
each new best. Penalty weights for load, time-window and distance violations are
adapted online. Tune any of this through [ils_params()].
