---
title: "Problem variants"
description: >
  A tour of every VRP variant vrpr supports, with worked examples and plots.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.4,
  fig.align = "center",
  dpi = 120,
  out.width = "85%"
)
```

```{r}
#| label: setup
#| message: false
library(vrpr)
```

`vrpr` exposes every variant of the classic vehicle routing problem through the
same tidy interface: you describe depots, clients and vehicle types, and the
solver figures out the rest. This article walks through each one.

Throughout, we stop the solver after a fixed number of iterations to keep the
article reproducible; in practice you would usually pass a time budget such as
`max_runtime(10)`.

## Capacitated VRP (CVRP)

The base case: clients have a `demand`, vehicles a `capacity`, and we minimise
total travel distance.

```{r}
#| label: cvrp
set.seed(42)
clients <- tibble::tibble(
  x = round(runif(24, -60, 60)),
  y = round(runif(24, -60, 60)),
  demand = sample(5:18, 24, replace = TRUE)
)

cvrp <- vrp_model() |>
  add_depot(0, 0) |>
  add_clients(clients) |>
  add_vehicle_type(num_available = 6, capacity = 70)

res <- vrp_solve(cvrp, stop = max_iterations(800), seed = 1, display = FALSE)
res
plot(res)
```

## Time windows (VRPTW)

Add `tw_early`, `tw_late` and `service` columns to make each client serviceable
only within a window. The solver respects the windows, and `routes()` reports the
`start_service` and `wait` time of every visit.

```{r}
#| label: vrptw
tw_clients <- tibble::tibble(
  x        = c(10, 20, 30, 40, 50, 60),
  y        = 0,
  demand   = 10,
  tw_early = c(0, 30, 60, 90, 120, 150),
  tw_late  = c(50, 80, 110, 140, 170, 200),
  service  = 10
)

vrptw <- vrp_model() |>
  add_depot(0, 0, tw_early = 0, tw_late = 500) |>
  add_clients(tw_clients) |>
  add_vehicle_type(num_available = 2, capacity = 60, tw_early = 0, tw_late = 500)

res_tw <- vrp_solve(vrptw, stop = max_iterations(500), seed = 1, display = FALSE)
routes(res_tw)[, c("route_id", "client", "start_service", "wait")]
```

Notice that the solver may *delay departure* so that no time is wasted waiting:
here every visit starts exactly within its window with zero waiting.

## Heterogeneous fleet

Call `add_vehicle_type()` more than once for a fleet of different vehicles --
different capacities, fixed costs, distance costs or shifts. Below, an expensive
type costs ten times more per unit distance than a cheap one of the same
capacity; the solver only uses the cheap type.

```{r}
#| label: heterogeneous
het <- vrp_model() |>
  add_depot(0, 0) |>
  add_clients(clients) |>
  add_vehicle_type(num_available = 4, capacity = 70, unit_distance_cost = 1) |>
  add_vehicle_type(num_available = 4, capacity = 70, unit_distance_cost = 10)

res_het <- vrp_solve(het, stop = max_iterations(800), seed = 1, display = FALSE)
table(`vehicle type` = routes(res_het)$vehicle_type)
```

## Multiple depots (MDVRP)

Add several depots and base each vehicle type at one of them with
`add_vehicle_type(depot = i)`. The `routes()` output gains a `depot` column, and
the solver assigns each client to the nearest depot's vehicles.

```{r}
#| label: mdvrp
mdvrp <- vrp_model() |>
  add_depot(x = -50, y = 0) |>
  add_depot(x =  50, y = 0) |>
  add_clients(tibble::tibble(
    x = c(-55, -45, -50, -48, 55, 45, 50, 52),
    y = c(5, -5, 10, -8, 5, -5, 8, -6),
    demand = 10
  )) |>
  add_vehicle_type(num_available = 3, capacity = 50, depot = 1) |>
  add_vehicle_type(num_available = 3, capacity = 50, depot = 2)

res_md <- vrp_solve(mdvrp, stop = max_iterations(500), seed = 1, display = FALSE)
plot(res_md)
```

## Prize-collecting

Mark clients as optional with `required = FALSE` and give them a `prize`. The
solver visits an optional client only when its prize offsets the routing cost;
`unvisited_clients()` lists those left out (drawn as hollow circles by `plot()`).

```{r}
#| label: prize
pc <- vrp_model() |>
  add_depot(0, 0) |>
  add_clients(tibble::tibble(
    x = c(5, -5, 0, 100, 100, 100),
    y = c(5, -5, 8, 10, 0, -10),
    demand = 10,
    required = c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE),
    prize = c(0, 0, 0, 5, 500, 5)
  )) |>
  add_vehicle_type(num_available = 4, capacity = 50)

res_pc <- vrp_solve(pc, stop = max_iterations(500), seed = 1, display = FALSE)
unvisited_clients(res_pc)
plot(res_pc)
```

`add_client_group()` goes further: it defines a set of mutually exclusive
alternatives, of which at most one (or exactly one, if `required = TRUE`) is
visited.

## Pickup & delivery / backhaul

Give clients a `pickup` amount in addition to their `demand` (delivery). The
collected load accumulates along the route and counts toward the vehicle's
capacity, modelling simultaneous pickup and delivery (and backhaul).

```{r}
#| label: pickup
pd <- vrp_model() |>
  add_depot(0, 0) |>
  add_clients(tibble::tibble(
    x = c(20, 40, -20, -40), y = 0,
    demand = 20, pickup = 30
  )) |>
  add_vehicle_type(num_available = 3, capacity = 60)

res_pd <- vrp_solve(pd, stop = max_iterations(300), seed = 1, display = FALSE)
res_pd$is_feasible
```

## Multi-trip

Allow a vehicle to return to a depot mid-route to reload, with
`add_vehicle_type(reload_depots = i, max_reloads = k)`. A single vehicle can then
serve far more than its capacity in one shift; `summary()$num_trips` counts the
trips.

```{r}
#| label: multitrip
mt <- vrp_model() |>
  add_depot(0, 0) |>
  add_clients(tibble::tibble(x = c(10, -10, 0, 5), y = c(0, 0, 10, -10), demand = 30)) |>
  add_vehicle_type(num_available = 1, capacity = 50, reload_depots = 1, max_reloads = 10)

res_mt <- vrp_solve(mt, stop = max_iterations(500), seed = 1, display = FALSE)
summary(res_mt)[, c("cost", "num_routes", "num_trips")]
```

A single vehicle (capacity 50) serves four clients of demand 30 -- 120 in total --
by making several trips back to the depot.
